Continuous gravitational waves in the lab: recovering audio signals with a table-top~optical microphone

### 2021-09-26 paper/ifo-appendix.tex
working on editors monor comments and adding reminders for larger rephrasings
comments from Andrew on appendix

\section{Detecting a sinusoidal signal in Gaussian noise}
\label{app:sinusoid_likelihood}
%\jam{[to do! from Andrew: ``Write a short appendix proving that maximizing the likelihood exp(-n||n/2) with respect to A given a signal A cos (wt + phi) is the Fourier transform. Classic sig proc result useful for teachers to know.'']}
In this appendix, we demonstrate that the modulus of the Fourier transform is an appropriate detection statistic when searching for a sinusoidal signal in Gaussian noise. 
We describe the data as
\begin{equation}
x(t) = s(t) + n(t)\,, 
\label{eqn:GNdata}
\end{equation}
where $s(t)$ and $n(t)$ are the signal and noise, respectively.
The signal takes the form
\begin{equation}
s(t) = A \cos\left[{\omega t + \phi}\right]\,,
\label{eqn:GNmodel}
\end{equation}
where $A$, $\omega$, and $\phi$ are the amplitude, angular frequency and phase of the signal, respectively. 
The noise $n(t)$ is a fluctuating zero-mean time series with the following property: if we define an inner product between two time series $u(t)$ and $v(t)$ as 
\begin{equation}
\ip{u}{v} = \frac{1}{T} \int_0^T \mathrm{d} t \, u(t) v(t)\,,
\label{eqn:ipuv}
\end{equation}
where $T$ is the total time of the observation, then the likelihood $\mathcal{L}$ of measuring the noise-related value $\ip{n}{n}$ is given by 
\begin{equation}
\mathcal{L} = \exp\left( -\frac{1}{2}\ip{n}{n}\right)\,.
\label{eqn:ipnn}
\end{equation}
Equations~\ref{eqn:ipuv} and~\ref{eqn:ipnn} define Gaussian noise through the fundamental measurement of $\ip{n}{n}$.

@@

Two important points must be made about the above procedure. 
(i) Fundamentally the goal is to maximize $\mathcal{L}$ in Eqn.~\ref{eqn:likeOne} by varying the signal.
For the special case of the Gaussian likelihood (Eqn.~\ref{eqn:ipnn}), which peaks at $\ip{n}{n}= 0$, this is equivalent to minimizing the distance between $x(t)$ and $s(t)$ defined through Eqn.~\ref{eqn:ipuv}.
In general, however, it is not equivalent to minimizing the distance between $x(t)$ and $s(t)$, for example if $\mathcal{L}$ peaks at $\ip{n}{n} \neq 0$, or if $\mathcal{L}$ has multiple maxima.
(ii) The maximum likelihood $\mathcal{L}_{\rm max}$ in Eqn.~\ref{eqn:maxLogLike} defines the detection statistic. 
When its value exceeds a threshold (chosen freely by the analyst) at some value of $\omega$, a signal is deemed to have been deteccted at that value of $\omega$.
Therefore the specific functional form of Eqn.~\ref{eqn:maxLogLike} matters, which is a second reason why one must start from Eqn.~\ref{eqn:ipnn} rather than Eqn.~\ref{eqn:likeOne}, in addition to reason (i).

@@

%\subsection{Pseudocode}
Here, we present some pseudocode (below) of the implementation used in Section~\ref{sec:viterbi}. We use Fourier amplitudes, normalized between $(0, 1)$ by dividing by the maximum value in the grid, as multiplicative weights. To avoid numerical underflow we take the logarithm and consider them instead as additive weights.

>>>>

\section{Detecting a sinusoidal signal in Gaussian noise}
\label{app:sinusoid_likelihood}

%\jam{[to do! from Andrew: ``Write a short appendix proving that maximizing the likelihood exp(-n||n/2) with respect to A given a signal A cos (wt + phi) is the Fourier transform. Classic sig proc result useful for teachers to know.'']}
In this appendix, we demonstrate that the modulus of the Fourier transform is an appropriate detection statistic when searching for a sinusoidal signal in Gaussian noise. 
We describe the data as
\begin{equation}
x(t) = s(t) + n(t)\,, 
\label{eqn:GNdata}
\end{equation}
where $s(t)$ and $n(t)$ are the signal and noise, respectively.
The signal takes the form
\begin{equation}
s(t) = A \cos\left[{\omega t + \phi}\right]\,,
\label{eqn:GNmodel}
\end{equation}
where $A$, $\omega$, and $\phi$ are the amplitude, angular frequency and phase of the signal, respectively. 
The noise $n(t)$ is a fluctuating zero-mean time series with the following property: if we define an inner product between two arbitrary time series $u(t)$ and $v(t)$ as 
\begin{equation}
\ip{u}{v} = \frac{1}{T} \int_0^T \mathrm{d} t \, u(t) v(t)\,,
\label{eqn:ipuv}
\end{equation}
where $T$ is the total time of the observation, then the likelihood $\mathcal{L}$ of measuring the noise-noise product $\ip{n}{n}$ is given by 
\begin{equation}
\mathcal{L} = \exp\left( -\frac{1}{2}\ip{n}{n}\right)\,.
\label{eqn:ipnn}
\end{equation}
Equations~\ref{eqn:ipuv} and~\ref{eqn:ipnn} define what it meand for noise to be Gaussian through the fundamental measurement of $\ip{n}{n}$.

@@

Two important points must be made about the above procedure. 
(i) Fundamentally the goal is to maximize $\mathcal{L}$ in Eqn.~\ref{eqn:likeOne} by varying $s(t)$ through $A$.
For the special case of the Gaussian likelihood (Eqn.~\ref{eqn:ipnn}), which peaks at $\ip{n}{n}= 0$, this is equivalent to minimizing the distance between $x(t)$ and $s(t)$ as evident in Eqn.~\ref{eqn:ipuv}.
In general, however, minimizing the distance between $x(t)$ and $s(t)$ is not equivalent always to the fundamental goal of maximising $\mathcal{L}$, for example if $\mathcal{L}$ peaks at $\ip{n}{n} \neq 0$, or if $\mathcal{L}$ has multiple maxima.
(ii) The maximum likelihood $\mathcal{L}_{\rm max}$ in Eqn.~\ref{eqn:maxLogLike} (or equivalently its logarithm) defines the detection statistic. 
When its value exceeds a threshold (chosen freely by the analyst) at some value of $\omega$, a signal is deemed to have been deteccted at that value of $\omega$.
Therefore the specific functional form of Eqn.~\ref{eqn:maxLogLike} matters, which is a second reason why one must start from Eqn.~\ref{eqn:ipnn} rather than Eqn.~\ref{eqn:likeOne}, in addition to reason (i).

@@

%\subsection{Pseudocode}
Here, we present some pseudocode (below) of the implementation used in Section~\ref{sec:viterbi}. We use Fourier amplitudes, normalized between $(0, 1)$ by dividing by the maximum value in the grid, as multiplicative weights. 
To avoid numerical underflow we take the logarithm of the Fourier amplitudes, which we can equivalently use as additive weights.
